---
title: "Generating a Count Matrix with featureCounts"
author: W. Evan Johnson
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: "flatly"
editor_options:
  chunk_output_type: console
---

## From alignments to counts

In the previous workshop we downloaded the HIV/TB FASTQ files and aligned
them to the human genome with **STAR**, producing one sorted BAM file per
sample. To analyze the data we need to summarize those alignments into a
**gene-by-sample count matrix**.

We use **featureCounts** (part of the [Subread](https://subread.sourceforge.net/)
package) to count, for each gene, how many sequencing reads align to it in
each sample.

## Set up Subread (featureCounts)

Subread is pre-installed in the bootcamp community space. The helper script
below adds `featureCounts` to your `PATH` (by appending to `~/.bashrc`) so it
is available every time you log in. It is safe to run more than once -- it
only adds the path if it is not already present.

Save this as `get_featureCounts_subread.sh` and run it with
`bash get_featureCounts_subread.sh`:

```{bash, eval=F}
#!/bin/bash

# Path to be added
SUBREAD_PATH="/projects/community/classes/bmihai_camp/apps/subread-2.0.3-Linux-x86_64/bin"

# Check if the path is already in .bashrc
if ! grep -Fxq "export PATH=\$PATH:$SUBREAD_PATH" ~/.bashrc; then
    echo "" >> ~/.bashrc
    echo "# Add Subread (featureCounts) to PATH" >> ~/.bashrc
    echo "export PATH=\$PATH:$SUBREAD_PATH" >> ~/.bashrc
    echo "Subread path added to ~/.bashrc."
else
    echo "Subread path already present in ~/.bashrc."
fi

source ~/.bashrc
```

Confirm it worked:

```{bash, eval=F}
featureCounts -v
```

## Run featureCounts

This command counts reads for **all** sorted BAM files at once and writes a
single counts table. Save it as `run_featureCounts.sh`:

```{bash, eval=F}
#!/bin/bash

featureCounts -g gene_name -t gene -O --fraction -T 32  \
  -a /projects/community/classes/bmihai_camp/STAR_index_human/gencode.v47.primary_assembly.annotation.gtf \
  -o /scratch/$USER/hivtb_data/counts_table.txt \
  $(printf "%s " /scratch/$USER/hivtb_data/aligned/*.sorted.bam)
```

### What the options mean

| Option | Meaning |
|--------|---------|
| `-g gene_name` | Group features by the `gene_name` attribute in the GTF (column names become gene symbols) |
| `-t gene` | Count reads over features of type `gene` |
| `-O` | Assign reads that overlap more than one feature to **all** of them |
| `--fraction` | Split multi-overlapping/multi-mapping reads fractionally instead of counting them whole |
| `-T 32` | Use 32 threads |
| `-a ...gtf` | Path to the GENCODE annotation (GTF) used for the STAR index |
| `-o ...counts_table.txt` | Output count matrix (plus a `.summary` file) |
| `$(printf ... *.sorted.bam)` | Expands to the list of all sorted BAM files to quantify |

## Run it as a SLURM batch job

`featureCounts` is multi-threaded, so submit it as a single batch job rather
than running it on the login node. Save the following and submit with
`sbatch`:

```{bash, eval=F}
#!/bin/bash
#SBATCH --job-name=hivtb_featurecounts
#SBATCH --output=featurecounts.out
#SBATCH --cpus-per-task=32
#SBATCH --mem=32G
#SBATCH --time=02:00:00
#SBATCH --reservation=bmihai-bootcamp-june2026
#SBATCH --partition=main-redhat

cd /scratch/$USER/hivtb_data

# Make featureCounts available (or run get_featureCounts_subread.sh first)
export PATH=$PATH:/projects/community/classes/bmihai_camp/apps/subread-2.0.3-Linux-x86_64/bin

featureCounts -g gene_name -t gene -O --fraction -T 32 \
  -a /projects/community/classes/bmihai_camp/STAR_index_human/gencode.v47.primary_assembly.annotation.gtf \
  -o /scratch/$USER/hivtb_data/counts_table.txt \
  $(printf "%s " /scratch/$USER/hivtb_data/aligned/*.sorted.bam)
```

## Inspect and clean the output

`featureCounts` writes two files:

- `counts_table.txt` -- the count matrix
- `counts_table.txt.summary` -- per-sample assignment statistics

The matrix begins with a header comment line and several annotation columns
(`Geneid`, `Chr`, `Start`, `End`, `Strand`, `Length`) before the per-sample
count columns. The BAM file path is used as each sample's column name, so the
columns usually need to be tidied (drop the annotation columns and shorten the
sample names) to produce the `features_combined.txt` used in the downstream
analysis.

```{bash, eval=F}
# Peek at the first few rows/columns
head counts_table.txt

# Check how many reads were assigned per sample
column -t counts_table.txt.summary | head
```

A quick tidy in R:

```{r, eval=F}
counts <- read.table("counts_table.txt", header = TRUE, row.names = 1,
                     sep = "\t", check.names = FALSE)

## Drop the annotation columns; keep only the per-sample count columns
counts <- counts[, -(1:5)]

## Shorten BAM paths to clean sample names (edit the pattern to match yours)
colnames(counts) <- sub(".*/", "", colnames(counts))
colnames(counts) <- sub(".sorted.bam$", "", colnames(counts))

write.table(counts, "features_combined.txt", sep = "\t", quote = FALSE)
```

The resulting `features_combined.txt` is the starting point for the
dimension-reduction and differential-expression analyses.

## Session Info
```{r session}
sessionInfo()
```
